Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-12-07 manuscript/agujournaltemplate.tex
stash full paper
pas text edits

\begin{keypoints}
\item Geometric models to predict lake depth, which require in-lake slope, assume that nearshore land slope is a good proxy for in-lake slope.
\item Using data from thousands of lakes, we show that nearshore land slope is a poor proxy for in-lake slope and increases prediction error.
\item Systematic overprediction of depth in concave and reservoir lakes highlights the potential for improvement in future depth prediction efforts.
\end{keypoints}

@@

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes. Given these knowledge gaps, we asked three research questions: 1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a lake compared to the distance to the deepest point of a lake? 2) How does the use of proxies for lake geometry affect lake depth prediction error? 3) How does lake cross-section shape (i.e. concave versus convex) and lake type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes from bathymetric map data and supplemented this data with classification estimates of whether lakes are reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Covariates used in offset modeling included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table S1).

By definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake). Therefore, we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in lakes with differing cross-section shape such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted (Figure S1). Finally, we expect that depth predictions themselves will be strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \cite{oliver2016prediction}.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on satellite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow. This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.
% We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results.

@@

Although proxy distance to lake center was often larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to lake center is an upper bound on distance to the deepest point of lakes. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure \ref{fig2}). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of lakes and distance to the deepest point means that it is possible to compute a "correction factor" to convert between the two measures (See best-fit equations in Figure \ref{fig2}).

@@

Model fit and prediction error differed depending on the technique used to calculate in-lake and nearshore geometry metrics (Table S2). We found that the best model fit and lowest model error occurred when in-lake slope was calculated as the average point-wise slope of all points at maximum lake depth rather than a single point of maximum depth. However, given the small difference in the fit of models using either of these techniques and the significant cost in terms of computational load and complexity, we limit our discussion hereafter to the simpler case involving only a single deepest point.

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S7). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S7). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Equation \ref{eq1} (Figure S7).

>>>>

\begin{keypoints}
\item Geometric models to predict lake depth, which require in-lake slope, assume that nearshore land slope is a good proxy for in-lake slope.
\item Using data from thousands of lakes, we show that nearshore land slope is a poor proxy for in-lake slope and increases prediction error.
\item Systematic overprediction of depth in concave and reservoir lakes highlights potential areas for improved depth prediction.
\end{keypoints}

@@

As a result of this limited testing, we lack knowledge on both the predictive performance of geometric models, the effect of proxies on depth prediction, and whether depth predictions are more sensitive to measurement errors in the horizontal dimension (i.e. distance to the deepest point of the lake) or measurement errors in the vertical dimension (i.e. in-lake slope). Additionally it is unclear whether model prediction error is related to differences in lake type such those with different cross-section shapes or those classified as reservoirs versus natural lakes. To address these knowledge gaps, we asked three research questions: 1) How representative is nearshore land slope of in-lake slope; and how representative is the distance to the center of a lake compared to the distance to the deepest point of a lake? 2) How does the use of proxies for lake geometry affect lake depth prediction error? 3) How does lake cross-section shape (i.e. concave versus convex) and lake type (i.e. natural lake vs reservoir) affect depth prediction error? To answer these questions, we extracted maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, cross-section shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes from bathymetric map data. We supplemented these geometry measures with data classifying waterbodies as reservoirs or natural lakes. We used this data to compute geometric depth estimates (Equation \ref{eq1}) and prediction "offsets" to these estimates using the random forest algorithm (Equation \ref{eq3}). Such offsets are model quantities which minimize differences between observed and predicted depth as a function of covariates. In our case, covariates included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table S1).

By definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake). Therefore, we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \cite{lanza1985interactions}. In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in lakes with differing cross-section shape such that the depth of bowl-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted (Figure S1). Finally, we expect that depth predictions themselves will be strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \cite{oliver2016prediction}.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on satellite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow (S. Polus, personal communication, 2020). This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.
% We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results.

@@

Although proxy distance to lake center was often larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to lake center is an upper bound on distance to the deepest point of lakes. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure \ref{fig2}). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of lakes and distance to the deepest point means that it is possible to convert between the two measures in subsequent analyses (See best-fit equations in Figure \ref{fig2}).

@@

Offset model fit and prediction error differed depending on the technique used to calculate in-lake and nearshore geometry metrics (Table S2). We found that the best model fit and lowest model error occurred when in-lake slope was calculated as the average point-wise slope of all points at maximum lake depth rather than a single point of maximum depth. However, given the small difference in the fit of models using either of these techniques and the significant cost in terms of computational load and complexity, we limit our discussion hereafter to the simpler case involving only a single deepest point.

@@

The most important covariates in these models were those relating to spatial location, lake area, and perimeter (Figure S7). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit. The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Equation \ref{eq1} (Figure S7).
